Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
questions are now TODO
edits to intro

\emph{[Alternate introduction, trying to tie the points more closely
to the content of the paper:\\
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
The performance of our new implementation is comparable to the
performance of~C++.
\item
As expected, higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
Haskell is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time (\secref{penumbra}).
We~found it difficult to tell which is which in advance.
\item
It is possible to use Haskell successfully for research even with a
very modest background in functional programming. 
\item
The only significant impediment that Haskell pas presented to our
research has been the difficulty of understanding and tuning the
performance of Haskell programs.
\end{itemize}
]}

@@

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
We~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
\remark{Two questions: (1)~Why isn't the function called \texttt{vee}?
(2)~Do we prefer \texttt{minimum (map from [Mat, Ins, Del])}?
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

\subsection{Performance}

@@

\section{Our previous experience compared}

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
Our~prior experience therefore includes building,
maintaining, and extending research tools written in~C++.
One~such tool is~Matt,
a~structural protein aligner~\cite{Menke:2008wu}.
Matt comprises about 12,000 lines of~C++.
It uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
It relies on no external libraries beyond \texttt{zlib},
but contains 1,000 lines of protein-structure code that ought
to be in a library.

@@

Like many researchers, we~and our colleagues often write code hastily.
Hastily written C++ code can get a job done, 
but we have found it expensive in the long run.
Our most painful experience was with an implementation of
\textit{multidimensional}
dynamic programming which was hastily added to the SMURF homology detector
\cite{Menke:2010ti}.
The implementation took 9,000~lines, including a single function of
over 700 lines.
When we added simulated evolution to this code, a relatively minor
enhancement \cite{Daniels:2012}, 
just understanding the existing code took several months.
Part of SMURF's code includes the 1,000 lines of protein-structure code
copied from the previous Matt software.

>>>>

\emph{[Alternate introduction, trying to tie the points more closely
to the content of the paper:\\
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
The performance of our new implementation is comparable to the
performance of~C++ (\secref{perf}).
\item
As expected, higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
Haskell is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time (\secref{penumbra}).
We~found it difficult to tell which is which in advance.
\item
It is possible to use Haskell successfully for research even with a
very modest background in functional programming. We found our Haskell
implementation to be more flexible about adding new research ideas
than our past C++ code (\secref{comparo}).
\item
The only significant impediment that Haskell has presented to our
research has been the difficulty of understanding and tuning the
performance of Haskell programs.
\end{itemize}
]}

@@

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
We~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
\remark{TODO: (1)~Call the inner function \texttt{vee}.
(2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

\subsection{Performance}
\seclabel{perf}

@@

\section{Our previous experience compared}
\seclabel{comparo}

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tool or library it uses is \texttt{zlib}.
Its initial development took \emph{[HOW LONG?]}.
\item
SMURF
\cite{Menke:2010ti} is used \emph{[FOR WHAT?]}
It~comprises about \emph{[HOW MANY?]} lines of~C++.
Its initial development took \emph{[HOW LONG?]}.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about \emph{[HOW MANY?]} lines of~Haskell
Its initial development took about three months of part-time work.
\remark{Loss of parallel structure---do we distinguish part time from
full time?}
\end{itemize}

It uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
It relies on no external libraries beyond \texttt{zlib},
but contains 1,000 lines of protein-structure code that ought
to be in a library.

@@

Like many researchers, we~and our colleagues often write code hastily.
Hastily written C++ code can get a job done, 
but we have found it expensive in the long run.
Our most painful experience was with an implementation of
\textit{multidimensional}
dynamic programming which was hastily added to the SMURF homology detector

The implementation took 9,000~lines, including a single function of
over 700 lines.
When we added simulated evolution to this code, a relatively minor
enhancement \cite{Daniels:2012}, 
just understanding the existing code took several months.
Part of SMURF's code includes the 1,000 lines of protein-structure code
copied from the previous Matt software.
